---
title: "Outlier"
output: html_document
---

First, use initialExploration.Rmd to prepare data
-imputes missing values with column mean
-remember allMesonet is the variable to use
Second, normalize data
```{r}
normFARM <- scale(class.FARM[,c(-1,-2,-3,-4,-5,-6,-7)])
normGRHM <- scale(class.GRHM[,c(-1,-2,-3,-4,-5,-6,-7)])
normHTFD <- scale(class.HTFD[,c(-1,-2,-3,-4,-5,-6,-7)])
normMRRY <- scale(class.MRRY[,c(-1,-2,-3,-4,-5,-6,-7)])
normPGHL <- scale(class.PGHL[,c(-1,-2,-3,-4,-5,-6,-7)])
normRSVL <- scale(class.RSVL[,c(-1,-2,-3,-4,-5,-6,-7)])
normAllMeso <- scale(allMesonet[,c(-1,-2,-3,-4,-5,-6,-7)])

normFARM <- scale(class.FARM[,c(-1,-2,-3,-4,-5,-6,-7)])
normGRHM <- scale(class.GRHM[,c(-1,-2,-3,-4,-5,-6,-7)])
normHTFD <- scale(class.HTFD[,c(-1,-2,-3,-4,-5,-6,-7)])
normMRRY <- scale(class.MRRY[,c(-1,-2,-3,-4,-5,-6,-7)])
normPGHL <- scale(class.PGHL[,c(-1,-2,-3,-4,-5,-6,-7)])
normRSVL <- scale(class.RSVL[,c(-1,-2,-3,-4,-5,-6,-7)])
normAllMeso <- scale(allMesonet[,c(-1,-2,-3,-4,-5,-6,-7)])
```

PCA
```{r}
pcaFARM <- prcomp(normFARM, 
                  center = TRUE,
                  scale. = TRUE)
pcaGRHM <- prcomp(normGRHM, 
                  center = TRUE,
                  scale. = TRUE)
pcaHTFD <- prcomp(normHTFD, 
                  center = TRUE,
                  scale. = TRUE)
pcaMRRY <- prcomp(normMRRY, 
                  center = TRUE,
                  scale. = TRUE)
pcaPGHL <- prcomp(normPGHL, 
                  center = TRUE,
                  scale. = TRUE)
pcaRSVL <- prcomp(normRSVL, 
                  center = TRUE,
                  scale. = TRUE)
pcaAllMeso <- prcomp(normAllMeso, 
                  center = TRUE,
                  scale. = TRUE)
```
Print those results
```{r}
print(pcaFARM)
print(pcaGRHM)
print(pcaHTFD)
print(pcaMRRY)
print(pcaPGHL)
print(pcaRSVL)
print(pcaAllMeso)

plot(pcaFARM, type = "l")
plot(pcaGRHM, type = "l")
plot(pcaHTFD, type = "l")
plot(pcaMRRY, type = "l")
plot(pcaPGHL, type = "l")
plot(pcaRSVL, type = "l")
plot(pcaAllMeso, type = "l")
```

Now do PCA?
```{r}
library(caret)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
trans <- preProcess(normAllMeso, 
                    method=c("BoxCox", "center",
                             "scale", "pca"))
PC = predict(trans, normAllMeso)
head(PC)
```

create random sample sizes of locations
```{r}
splace <- sample(nrow(normFARM), 0.5*nrow(normFARM))
sampleFARM <- normFARM[splace,]
splace <- sample(nrow(normGRHM), 0.5*nrow(normGRHM))
sampleGRHM <- normGRHM[splace,]
splace <- sample(nrow(normHTFD), 0.5*nrow(normHTFD))
sampleHTFD <- normHTFD[splace,]
splace <- sample(nrow(normMRRY), 0.5*nrow(normMRRY))
sampleMRRY <- normMRRY[splace,]
splace <- sample(nrow(normPGHL), 0.5*nrow(normPGHL))
samplePGHL <- normPGHL[splace,]
splace <- sample(nrow(normRSVL), 0.5*nrow(normRSVL))
sampleRSVL <- normRSVL[splace,]
splace <- sample(nrow(normAllMeso), nrow(normAllMeso))
sampleMeso <- normAllMeso[splace,]
```

Third, use outlier libraries to find outliers per location
```{r}
library(DMwR2)
#normFARM
#normGRHM
#normHTFD
#normMRRY
#normPGHL
#normRSVL
#normAllMeso

ptm <- proc.time()
outlier.scores.farm <- lofactor(normFARM, k=5)
outlier.scores.farm.o <- outlier.scores.farm[which(is.nan(outlier.scores.farm) == 0)]
outlier.scores.grhm <- lofactor(normGRHM, k=5)
outlier.scores.grhm.o <- outlier.scores.grhm[which(is.nan(outlier.scores.grhm) == 0)]
outlier.scores.htfd <- lofactor(normHTFD, k=5)
outlier.scores.htfd.o <- outlier.scores.htfd[which(is.nan(outlier.scores.htfd) == 0)]
outlier.scores.mrry <- lofactor(normMRRY, k=5)
outlier.scores.mrry.o <- outlier.scores.mrry[which(is.nan(outlier.scores.mrry) == 0)]
outlier.scores.pghl <- lofactor(normPGHL, k=5)
outlier.scores.pghl.o <- outlier.scores.pghl[which(is.nan(outlier.scores.pghl) == 0)]
outlier.scores.rsvl <- lofactor(normRSVL, k=5)
outlier.scores.rsvl.o <- outlier.scores.rsvl[which(is.nan(outlier.scores.rsvl) == 0)]
exclude <- which(is.nan(outlier.scores.rsvl.o) == 1)
exclude <- c(exclude, which(is.infinite(outlier.scores.rsvl.o) == 1))
outlier.scores.rsvl.o <- outlier.scores.rsvl.o[-exclude]
proc.time() - ptm

ptm <- proc.time()
outlier.scores.meso3 <- lofactor(normAllMeso, k=3)
proc.time() - ptm

exclude <- which(is.nan(outlier.scores.meso) == 1)
exclude <- c(exclude, which(is.infinite(outlier.scores.meso) == 1))
outlier.scores.meso.clean <- outlier.scores.meso[-exclude]
normAllMeso.clean <- normAllMeso[-exclude,]

plot(density(outlier.scores.meso))
plot(density(outlier.scores.meso.clean))
plot(density(outlier.scores.meso.clean), main="Local Outlier Factor Density Plot")

plot(density(outlier.scores.farm))
plot(density(outlier.scores.grhm))
plot(density(outlier.scores.htfd))
plot(density(outlier.scores.mrry))
plot(density(outlier.scores.pghl))
plot(density(outlier.scores.rsvl))
#who are outliers
#print(outliers)
```

Combine normalized data points and outlier scores. Then filter out extraneous scores.
```{r}
outlier.scores.all <- c(outlier.scores.farm, outlier.scores.grhm, 
                        outlier.scores.htfd, outlier.scores.mrry,
                        outlier.scores.pghl, outlier.scores.rsvl)

```

Find mean and standard deviation
```{r}
lofMean <- mean(outlier.scores.farm)
lofSD <- sd(outlier.scores.farm)
lofThresholdfarm <- length(which(outlier.scores.farm > (lofMean + lofSD*3)))
lofMean <- mean(outlier.scores.grhm)
lofSD <- sd(outlier.scores.grhm)
lofThresholdgrhm <- length(which(outlier.scores.grhm > (lofMean + lofSD*3)))
lofMean <- mean(outlier.scores.htfd)
lofSD <- sd(outlier.scores.htfd)
lofThresholdhtfd <- length(which(outlier.scores.htfd > (lofMean + lofSD*3)))
lofMean <- mean(outlier.scores.mrry)
lofSD <- sd(outlier.scores.mrry)
lofThresholdmrry <- length(which(outlier.scores.mrry > (lofMean + lofSD*3)))
lofMean <- mean(outlier.scores.pghl)
lofSD <- sd(outlier.scores.pghl)
lofThresholdpghl <- length(which(outlier.scores.pghl > (lofMean + lofSD*3)))
lofMean <- mean(outlier.scores.rsvl)
lofSD <- sd(outlier.scores.rsvl)
lofThresholdrsvl <- length(which(outlier.scores.rsvl > (lofMean + lofSD*3)))
#pick top 5% of outliers
outliersfarm <- order(outlier.scores.farm, decreasing=T)[1:lofThresholdfarm]
outliersgrhm <- order(outlier.scores.grhm, decreasing=T)[1:lofThresholdgrhm]
outliershtfd <- order(outlier.scores.htfd, decreasing=T)[1:lofThresholdhtfd]
outliersmrry <- order(outlier.scores.mrry, decreasing=T)[1:lofThresholdmrry]
outlierspghl <- order(outlier.scores.pghl, decreasing=T)[1:lofThresholdpghl]
outliersrsvl <- order(outlier.scores.rsvl, decreasing=T)[1:lofThresholdrsvl]
```

Outliers for all of mesonet
```{r}
lofMean <- mean(outlier.scores.meso.clean)
lofSD <- sd(outlier.scores.meso.clean)
rmExtraneousValues <- which(outlier.scores.meso.clean > (lofMean + lofSD*3))

#old calculations
lofThresholdmeso <- length(which(outlier.scores.meso.clean > (lofMean + lofSD*3)))
outliersmeso <- order(outlier.scores.meso.clean, decreasing=T)[1:lofThresholdmeso]
```

Plot Mesonet outliers
```{r}
n <- length(outlier.scores.meso.clean)
pch <- rep(".", n)
pch[rmExtraneousValues] <- "_"
col <- rep("black", n)
col[rmExtraneousValues] <- "red"
pairs(normAllMeso.clean, pch=pch, col=col)
```

Plot outliers
```{r}
exclude <- which(is.nan(outliershtfd) == 1)
exclude <- c(exclude, which(is.infinite(outliershtfd) == 1))

n <- nrow(normRSVL)
pch <- rep(".", n)
pch[outliersrsvl] <- "+"
col <- rep("black", n)
col[outliersrsvl] <- "red"
pairs(normRSVL, pch=pch, col=col)
```

Histogram code
```{r}
my.bin.width = 1
tairhist <- hist(normAllMeso.clean[,1], main = "TAIR Histogram")
summary(normAllMeso.clean[,1])
relhhist <- hist(normAllMeso.clean[,2], main = "RELH Histogram")
summary(normAllMeso.clean[,2])
tdpthist <- hist(normAllMeso.clean[,3], main = "TDPT Histogram")
summary(normAllMeso.clean[,3])
wspdhist <- hist(normAllMeso.clean[,4], main = "WSPD Histogram")
summary(normAllMeso.clean[,4])
wdirhist <- hist(normAllMeso.clean[,5], main = "WDIR Histogram")
summary(normAllMeso.clean[,5])
wsmxhist <- hist(normAllMeso.clean[,6], main = "WSMX Histogram")
summary(normAllMeso.clean[,6])
wdmxhist <- hist(normAllMeso.clean[,7], main = "WDMX Histogram")
summary(normAllMeso.clean[,7])
sradhist <- hist(normAllMeso.clean[,8], main = "SRAD Histogram")
summary(normAllMeso.clean[,8])
prcphist <- hist(normAllMeso.clean[,9], main = "PRCP Histogram")
summary(normAllMeso.clean[,9])
```

Thresholds
```{r}
length(which(normAllMeso.clean[,1] < -2.5))
length(which(normAllMeso.clean[,1] > 2.5))

length(which(normAllMeso.clean[,2] < -2.25))
length(which(normAllMeso.clean[,2] > 1.459))

length(which(normAllMeso.clean[,3] < -3))
length(which(normAllMeso.clean[,3] > 2))

length(which(normAllMeso.clean[,4] < -1))
length(which(normAllMeso.clean[,4] > 3))

length(which(normAllMeso.clean[,5] < -1.25))
length(which(normAllMeso.clean[,5] > 1.25))

length(which(normAllMeso.clean[,6] < -1))
length(which(normAllMeso.clean[,6] > 3))

length(which(normAllMeso.clean[,7] < -1.25))
length(which(normAllMeso.clean[,7] > 1.25))

length(which(normAllMeso.clean[,8] < -0.25))
length(which(normAllMeso.clean[,8] > 3))

length(which(normAllMeso.clean[,9] < -0.5))
length(which(normAllMeso.clean[,9] > 3))
```

