---
title: "Clustering"
output: html_document
---

Load data
Average out NAs
Scale data


```{r}
setwd("C:/Users/Owner/Desktop/DM_proj/Clustering")
load("clustering_env.RData")
FARM <- read.csv("FARM_Bowling-Green-5-S_Warren_Discrete.csv")
HTFD <- read.csv("HTFD_Hartford-3-E_Ohio_Discrete.csv")
PGHL <- read.csv("PGHL_Hopkinsville-6-N_Christian_Discrete.csv")
GRHM <- read.csv("GRHM_Henderson-5-E_Henderson_Discrete.csv")
MRRY <- read.csv("MRRY_Murray-1-W_Calloway_Discrete.csv")
RSVL <- read.csv("RSVL_Russellville-2-W_Logan_Discrete.csv")

#Trim unneeded columns
FARM$X <- NULL
GRHM$X <- NULL
HTFD$X <- NULL
MRRY$X <- NULL
PGHL$X <- NULL
RSVL$X <- NULL
```

Clean and preprocess data. Replace the NAs with the column average
```{r}
#all data in mesonet
mesonet <- list(FARM, GRHM, HTFD, MRRY, PGHL, RSVL)
#where newly processed data will be stored
newMeso <- list()
####Data Preprocessing
#Get NA values
j <- 0
for (DATA in mesonet){
  j <- j + 1
  na.DATA <- list()
  #make list of NAs for each column
  for(i in 1:ncol(DATA)){
    na.DATA <- c( na.DATA, list(which( is.na(DATA[,i]) ) ) )
    DATA[na.DATA[[i]],i]
  }
  #Print NA Values
  #for(i in 1:ncol(DATA)){
    #print(DATA[na.DATA[[i]],i])
  #}
  #Replace NA with column mean
  for(i in 1:ncol(DATA)){
    DATA[is.na(DATA[,i]), i] <- mean(DATA[,i], na.rm = TRUE)
  }
  #Print replaced NA Values
  #for(i in 1:ncol(DATA)){
    #print(DATA[na.DATA[[i]],i])
  #}
  #print(j)
  #store processed data in new variable
  newMeso[[j]] <- DATA
}
```

Check Preprocessed data for NAs (should no longer have NAs)
```{r}
for (DATA in newMeso){
  #Print NA Values
  na.DATA <- list()
  #loops through every column in Mesonet data structure
  for(i in 1:ncol(DATA)){
    na.DATA <- c( na.DATA, list(which( is.na(DATA[,i]) ) ) )
    DATA[na.DATA[[i]],i]
  }
  for(i in 1:ncol(DATA)){
    print(DATA[na.DATA[[i]],i])
  }
}
```

Combine all of the mesonet data into one table
```{r}
allMesonet <- rbind(newMeso[[1]],newMeso[[2]],newMeso[[3]],newMeso[[4]],newMeso[[5]],newMeso[[6]])
allMesonet$X <- NULL
```

convert nominal data to numeric data
```{r}
#create separate variables
allSTID <- allMesonet$STID
allNumSTID <- as.numeric(allSTID)
allMesonet$STID <- allNumSTID

```

Scale data.
The scale function finds the mean and standard deviation of each column.
It then takes each element in each column, subtracts it by the mean, then
divides it by standard deviation.
```{r}
#remove year because scale function says it's NA
allMesonet$Year <- NULL
scaledMesonet <- scale(allMesonet[,-1])
scaledMesonet <- cbind(scaledMesonet, allMesonet$STID)
colnames(scaledMesonet)[15] <- "STID"
#view the ranges of Zscores for each attribute
for(i in 1:ncol(scaledMesonet)) {
   print(range(scaledMesonet[,i]))
}
```


Create sample sizes
```{r}
set.seed(1234)
train <- array()
kfolds <- 10
i<-1
delta <- 0
randomIndices <- sample(nrow(scaledMesonet), nrow(scaledMesonet))
dataLength <- length(randomIndices)
for(i in 1:kfolds) {
  kRange1 <- ceiling(dataLength - dataLength*(i/kfolds))
  kRange2 <- dataLength - delta
  train[i] <- list(randomIndices[kRange1:kRange2])
  trainStr <- paste(c("meso.training.",i,".csv"), collapse = "")
  validateStr <- paste(c("meso.validate.",i,".csv"), collapse = "")
  write.csv(scaledMesonet[train[[i]],], file = trainStr)
  write.csv(scaledMesonet[-train[[i]],], file = validateStr)
  delta <- dataLength*(i/kfolds)
}
mesoSample <- scaledMesonet[train[[1]],]
mesoValidate <- scaledMesonet[-train[[1]],]

write.csv(scaledMesonet, file = "allMeso.csv")
```


